Detecting changes in dispersion in COVID-19 incidence time series using a negative binomial model

### 2023-09-30 noise_covid_v2.Rmd
updates

---
title: Detecting changes in dispersion in COVID-19 incidence time series using a negative binomial model
author: "Rachael Aber, Yanming Di, Ben Dalziel"
header-includes: \usepackage{setspace}\doublespacing
output:
  pdf_document: default
  html_document:
    df_print: paged
geometry: margin=2.54cm
fontsize: 12pt
---

@@

Evaluating infectious disease dynamics using only course-scale changes in incidence over time may fail to account for important parts of the social-biological context of transmission and control. For instance, the impact of non-pharmateutical interventions (NPIs; e.g., mask mandates) is often assessed in terms of changes in *mean* incidence, yet mask mandates are implemented and adhered to heterogeneously, so their effects may be context-dependent, leading to geographic and temporal variation in impact for the same policy. Similarly, the impacts of events/gatherings (such as national holidays) may result in changes in epidemic trajectories that differ based on context, and these changes may not be elucidated by simply examining aggregate mean changes in the US or even mean changes per county.

Implementing and evaluating targeted control strategies has typically required detailed data on individual-level variation in transmission rates, which are often unavailable. We present a method to identify shifts in population-level dispersion in incidence, so the impact of events or NPIs on superspreading dynamics can be assessed using freely available population-level data.

@@

Naively, the negative binomial distribution might accurately model a time series if there is a changing process mean: for example, if the mean of a Poisson distribution itself follows a gamma distribution, the resulting distribution is negative binomial (Cook 2009). Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion that all lead to overdispersion (Barron 1992). Improved understanding of the causes of overdispersion in incidence may provide insight into interactions between host population structure and contagion processes, contributing to general predictive understanding in a range of systems, and understanding the limits of predictability.Previously, efforts to quantify the processes underlying overdispersed incidence have included the study of underlying offspring distributions. Also, comparison of simulated and observed incidence time series has been used to estimate the role of environmental and demographic stochasticity in measles: across community sizes, demographic stochasticity of measles becomes more important in small human populations, where dynamics can’t be described as well simply by contact rate and birth rate (Grenfell et al. 2002). The final process potentially underlying overdispersed incidence, varying population effective reproduction number, has been studied extensively in investigations into seasonal forcing of population-level effective reproduction number.

**Using county-level SARS-CoV-2 time series from New York state, we investigate dispersion in case count time series and its relationship to a hypothesized changepoint. Specifically, we estimate and compare dispersion on either side of a putative changepoint in variability (Thanksgiving 2020), in order to test the hypothesis that dispersion changes significantly**. 
It was recently found that $k_t$, the time-varying transmission heterogeneity for COVID-19, decreased over time and was significantly associated with interventions to slow spread in Hong Kong (Adam et al. 2022). Since $k_t$ underlies incidence overdispersion, the method developed here will allow practitioners in epidemiology to explore whether local events result in a change in dispersion of cases, potentially modulated by changes in $k_t$. Key unanswered questions remain in developing methodology to recover incidence dispersion estimates and using these estimates to test the effect of gatherings and NPIs.

@@

We additionally use the full dataset to "scan" for variability breakpoints using a likelihood ratio test statistic.

@@

## Proposed hypothesis test

We tested the hypothesis that the incidence dispersion parameter changes at the Thanksgiving changepoint of interest: $\theta_1 \ne \theta_2$. Hypothesis tests were conducted to assess whether $\theta$ changes after Thanksgiving 2020. The Wald test was used, where the estimates of $\theta$ in different segments are considered independent. This approach is possible because we used IRLS estimates of $\theta$ - the standard error of $\theta$ was found utilizing the observed information matrix.

@@

$$Cov(\hat{\theta_1}, \hat{\theta_2}) = 0 $$
The form of the Wald test statistic is:

$$\frac{\hat{\theta_1} -\hat{\theta_2}}{SE(\hat{\theta_1} -\hat{\theta_2})}$$
$$ = \frac{\hat{\theta_1} -\hat{\theta_2}}{\sqrt{\widehat{Var(\hat{\theta_1})} + \widehat{Var(\hat{\theta_2})}}}$$
# Results

## Thanksgiving 2020: Wald test

For every county, the model was fit separately to either side of November 26th, 2020 in order to compute values of $\hat{\theta_1}$ and $\hat{\theta_2}$. A test of the two-sided alternative that $\theta$ before is different from $\theta$ after (different theta after Thanksgiving) was performed.

# Simulation

## Validity and power of the Wald test

@@

To ensure that the hypothesis test has adequate power, a simulation using unequal values of theta on either side of Thanksgiving was repeatedly run in order to ensure that the distribution of resulting p-values is heavily right-skewed. 
The resulting probabilities of rejection are tabulated below: the proportion of tests that reject is around 0.05 for those instances where the null is true (the probability of the normal variable having 0.05 of the probability density below is 0.05). Additionally, the proportion of tests that reject is high in instances where the alternative hypothesis is true. We also simulated data assuming that theta was a function of the mean of the process in order to assess the robustness of the test (Supplement).

![Type I Error Rate and Power - Wald Test](figures/pvals_sim_Wald_table.pdf)

Additionally, the relationship between the p-values resulting from the Wald test and the effect size is as expected.

![Wald Test p-values v. Theta Difference](figures/pvals_sim_Wald.pdf){width=30%, height=30%}

## Validity and power of the LRT

@@

@@

@@

>>>>

---
title: Detecting changes in dispersion in COVID-19 incidence time series using a negative binomial model
author: "Rachael Aber, Yanming Di, Ben Dalziel"
header-includes: \usepackage{setspace}\doublespacing
output:
  pdf_document: default
  html_document:
    df_print: paged
geometry: margin = 2.54cm
fontsize: 12pt
---

@@

Evaluating infectious disease dynamics using only course-scale changes in incidence over time may fail to account for important parts of the social-biological context of transmission and control. For instance, the impact of non-pharmateutical interventions (NPIs; e.g., mask mandates) is often assessed in terms of changes in *mean* incidence, yet mask mandates are implemented and adhered to heterogeneously, so their effects may be context-dependent, leading to geographic and temporal variation in impact for the same policy. Similarly, the impacts of gatherings (such as those on national holidays) may result in changes in epidemic trajectories that differ based on context, and these changes may not be elucidated by simply examining aggregate mean changes in the US or even mean changes per county.

We present a method to identify shifts in population-level dispersion in incidence, so the impact of events or NPIs on superspreading dynamics can be assessed using freely available population-level data. This method may also be useful in identifying geographical areas that would benefit from targeted control strategies to reduce superspreading dynamics. Implementing and evaluating targeted control strategies has typically required detailed data on individual-level variation in transmission rates, which are often unavailable.

@@

Improved understanding of the causes of overdispersion in incidence may provide insight into interactions between host population structure and contagion processes, contributing to general predictive understanding in a range of systems, and understanding the limits of predictability.Previously, efforts to quantify the processes underlying overdispersed incidence have included the study of underlying offspring distributions. Also, comparison of simulated and observed incidence time series has been used to estimate the role of environmental and demographic stochasticity in measles: across community sizes, demographic stochasticity of measles becomes more important in small human populations, where dynamics can’t be described as well simply by contact rate and birth rate (Grenfell et al. 2002). The final process potentially underlying overdispersed incidence, varying population effective reproduction number, has been studied extensively in investigations into seasonal forcing of population-level effective reproduction number.

## Negative Binomial Regression

The negative binomial distribution might accurately model a time series if there is a changing process mean: for example, if the mean of a Poisson distribution itself follows a gamma distribution, the resulting distribution is negative binomial (Cook 2009). Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion that all lead to overdispersion (Barron 1992).

**Using county-level SARS-CoV-2 time series from New York state, we investigate dispersion in case count time series (by fitting negative binomial models) and its relationship to a hypothesized changepoint. Specifically, we estimate and compare dispersion on either side of a putative changepoint in variability (Thanksgiving 2020), in order to test the hypothesis that dispersion changes significantly**. 
It was recently found that $k_t$, the time-varying transmission heterogeneity for COVID-19, decreased over time and was significantly associated with interventions to slow spread in Hong Kong (Adam et al. 2022). Since $k_t$ underlies incidence overdispersion, the method developed here will allow practitioners in epidemiology to explore whether local events result in a change in dispersion of cases, potentially modulated by changes in $k_t$. Key unanswered questions remain in developing methodology to recover incidence dispersion estimates and using these estimates to test the effect of gatherings and NPIs.

@@

We additionally use the full dataset to "scan" for variability breakpoints using a likelihood ratio test statistic and visualizations of a dispersion metric.

@@

## Proposed Likelihood Ratio Test

We tested the hypothesis that the incidence dispersion parameter changes at the Thanksgiving changepoint of interest: $\theta_1 \ne \theta_2$.

@@

@@

To ensure that the hypothesis test has adequate power, a simulation using unequal values of theta on either side of Thanksgiving was repeatedly run in order to ensure that the distribution of resulting p-values is heavily right-skewed.

The resulting probabilities of rejection are tabulated below: the proportion of tests that reject is around 0.05 for those instances where the null is true (the probability of the normal variable having 0.05 of the probability density below is 0.05). Additionally, the proportion of tests that reject is high in instances where the alternative hypothesis is true. We also simulated data assuming that theta was a function of the mean of the process in order to assess the robustness of the test (Supplement).

@@

# Results

@@

For every county, the negative binomial model was fit separately to either side of November 26th, 2020 in order to compute values of $\hat{\theta_1}$ and $\hat{\theta_2}$. A test of the two-sided alternative that $\theta$ before is different from $\theta$ after (different theta after Thanksgiving) was performed.

@@

## Less dispersion over time

In many US counties between 2020-09-27 and 2021-01-25, there is less burstiness (larger theta) in COVID-19 case counts over time. In other words, cases fluctuate less around a time-averaged trajectory later on, possibly indicating structure in the population with some groups experiencing different infection rates (overdispersed individual reproduction number), demographic/environmental stochasticity, or fast changes in propagation/reporting at the population level may have occurred earlier on. However, the trend of decreased burstiness over this period seems to be modulated by population size: among the largest population counties per state, those in the top half had an average theta change of approx. 28.95604 (less burstiness), while the small populations had an average change in theta of approx. -3.741856 (more burstiness). This means that in the large populations, dynamics became "smoother", but this didn't happen in the smaller populations, where more burstiness might be a result of cases "stuttering out". This is a key observation, as smoother and more predictable case count time series may require less surge capacity in hospitals.

We hypothesized that there would be an increase in burstiness (decrease in theta) around Thanksgiving 2020 due to people gathering and causing clusters of infection. However, we observed that in many counties, although there was more burstiness (a decrease in theta), it was a relatively small change, and not modulated by population size. The average difference between theta estimated in the window starting at 2020-11-26 and theta estimated in the window ending 2020-11-25 was -1.77944 in the top half and -1.104289 in the bottom half of population sizes. People may have avoided gathering due to personal preference, cultural norms, and other factors, contributing to less superspreading than anticipated

![Compare](./figures/compare.pdf){width=30%, height=30%}
